---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# patchwork

[![Travis-CI Build Status](https://travis-ci.org/thomasp85/patchwork.svg?branch=master)](https://travis-ci.org/thomasp85/patchwork)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/thomasp85/patchwork?branch=master&svg=true)](https://ci.appveyor.com/project/thomasp85/patchwork)
[![CRAN_Release_Badge](http://www.r-pkg.org/badges/version-ago/patchwork)](https://CRAN.R-project.org/package=patchwork)
[![CRAN_Download_Badge](http://cranlogs.r-pkg.org/badges/patchwork)](https://CRAN.R-project.org/package=patchwork)

The goal of `patchwork` is to make it ridiculously simple to combine separate
ggplots into the same graphic. As such it tries to solve the same problem as
`gridExtra::grid.arrange()` but using an API that incites exploration and 
iteration.

## Installation

You can install patchwork from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("thomasp85/patchwork")
```

## Example

The usage of `patchwork` is simple: just add plots together!

```{r example}
library(ggplot2)
library(patchwork)

p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))

p1 + p2
```

you are of course free to also add the plots together as part of the same 
plotting operation:

```{r}
ggplot(mtcars) +
  geom_point(aes(mpg, disp)) +
  ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))
```

layouts can be specified by adding a `plot_layout()` call to the assemble. This
lets you define the dimensions of the grid and how much space to allocate to the
different rows and columns

```{r}
p1 + p2 + plot_layout(ncol = 1, heights = c(3, 1))
```

This is all it does for now, but stay tuned as more functionality is added, such
as collapsing guides, and nesting plots...
