---
title: "Adding Annotation and Style"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Adding Annotation and Style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patchwork)
```

Sometimes you simply want to put multiple plots side by side and call it a day,
but often you want the end result to stand forth like a collective thing. To 
achieve that you would often add a title and other textual cues. This guide will
teach you how to do that.

As always, we start with some plots.

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

## Titles, subtitles and captions
One of the most needed things is to add descriptive text to your plot ensemble. 
To achieve this, you simply add it to your patchwork using `plot_annotation()`

```{r}
patchwork <- (p1 + p2) / p3
patchwork + plot_annotation(
  title = 'The surprising truth about mtcars',
  subtitle = 'These 3 plots will reveal yet-untold secrets about our beloved data-set',
  caption = 'Disclaimer: None of these plots are insightful'
)
```

It is important to note that plot annotations only have an effect on the 
top-level patchwork. Any annotation added to nested patchworks are (currently) 
lost. If you need to have annotations for a nested patchwork you'll need to wrap 
it in `wrap_elements()` with the side-effect that alignment no longer works.

## Tagging
Often, especially in scientific literature, multiple plots are collected in a
single figure and referred to by a tag. While such tags could be added manually,
it is much simpler to let patchwork handle it for you, using the auto-tagging
functionality. This is turned on by setting `tag_level` in `plot_annotation()`
to a value indicating the family of symbols to use for tagging: `'1'` for 
Arabic numerals, `'A'` for uppercase Latin letters, `'a'` for lowercase Latin
letters, `'I'` for uppercase Roman numerals, and `'i'` for lowercase Roman 
numerals.

```{r}
patchwork + plot_annotation(tag_levels = 'A')
```

Tags uses the theming of the plot they are applied to, so make sure they match 
up, or modify the theming of all plots using `&` (see the 
[Plot Assembly](assembly.html) guide).

```{r}
patchwork + 
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 8))
```

When the patchwork contains nested layouts the tagging will recurse into them by
default, but you can tell it to define a new tagging level with the `tag_level`
argument in `plot_layout()`. You can then provide multiple tag-types to 
`tag_levels` to define how subtagging should be enumerated.

```{r}
patchwork[[1]] <- patchwork[[1]] + plot_layout(tag_level = 'new')
patchwork + plot_annotation(tag_levels = c('A', '1'))
```

With `plot_annotation()`it is also possible to define separator, prefix, and 
suffix for the tag, but don't go overboard with it:

```{r}
patchwork + plot_annotation(tag_levels = c('A', '1'), tag_prefix = 'Fig. ',
                            tag_sep = '.', tag_suffix = ':')
```

The default ggplot2 theme puts the tag in its own row and column that will 
expand to fit. For longer tag text this will look weird, so it is better to 
place it on top of the plot region:

```{r}
patchwork + 
  plot_annotation(tag_levels = c('A', '1'), tag_prefix = 'Fig. ', tag_sep = '.', 
                  tag_suffix = ':') & 
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 8, hjust = 0, vjust = 0))
```

Lastly it is also possible to provide you own tag sequence instead of relying on
the build in ones. Passing a list of character vectors will do just that (note
that this can be mixed with the standard sequences):

```{r}
patchwork + 
  plot_annotation(tag_levels = list(c('#', '&'), '1'))
```

If you provide more plots than your custom sequence support the excess plots 
will get empty tags so make sure that there's enough.

## Styling the patchwork
Most of the style of the patchwork is made up by the themes of the individual 
patches. For a coherent look, don't mix videly different looks. Setting the 
background colour of a single plot to a different shade is an effective way to
highlight it, but e.g. different fonts or line widths will just look like a 
mess. The patchwork itself has a few elements itself that is succeptible to 
theming: A background, a margin, and title, subtitle & caption. The theme of
the patchwork is by default the default ggplot2 theme. It can be changed though,
in two different ways. The easiest is to simply use `&` with a theme element. 
This operator will add to the theme of all subplots as well as to the theme of
the patchwork itself:

```{r}
patchwork + 
  plot_annotation(title = 'The surprising truth about mtcars') & 
  theme(text = element_text('mono'))
```

If you need to address only the theme of the patchwork itself (e.g. for making
the patchwork title larger than the plot titles), it can be done with the 
`theme` argument in `plot_annotation()` (note that the use of one does not 
exclude the other):

```{r}
patchwork + 
  plot_annotation(title = 'The surprising truth about mtcars',
                  theme = theme(plot.title = element_text(size = 18))) & 
  theme(text = element_text('mono'))
```

## Want more?
Now you know how to annotate and style your patchwork. Read the other guides 
about [assembling](assembly.html) and [laying out](layout.html) patchwork to
master all of patchwork.
