---
title: "Plot Assembly"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Plot Assembly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patchwork)
```

Before plots can be laid out, they have to be assembled. Arguably one of
patchwork's biggest selling points is that it expands on the use of `+` in
ggplot2 to allow plots to be added together and composed, creating a natural
extension of the ggplot2 API. There is more to it than that though, and this
tutorial will teach you all about the different operators and functions
available for plot assembly.

As always, we'll start with a few well-known example plots:

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) +
  geom_point(aes(mpg, disp)) +
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) +
  geom_boxplot(aes(gear, disp, group = gear)) +
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) +
  geom_point(aes(hp, wt, colour = mpg)) +
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) +
  geom_bar(aes(gear)) +
  facet_wrap(~cyl) +
  ggtitle('Plot 4')
```

## Adding plots to the patchwork
At this point, it shouldn't come as a surprise that you can use `+` to add plots
together to form a patchwork. If it does, I'd suggest you start out with the
[Getting Started](../patchwork.html) guide, and come back once you've gone
through that. But, just to recap: you can add plots together, like so:

```{r}
p1 + p2
```

Using this approach, it is possible to assemble a number of plots, but that is
not the only thing that can be added. Other patchworks can be added, and will
create a new nested patchwork:

```{r}
patch <- p1 + p2
p3 + patch
```

### Adding tables
patchwork has deep support for working with tables powered by the gt package and
it provides you with the means to define what to align with the panel region as
well as whether to respect the table size when setting the size for the rows and
columns in the patchwork.

If you just want the standard settings you can add a gt table directly to a
plot

```{r}
p1 + gt::gt(mtcars[1:10, c('mpg', 'disp')])
```

but if you want to change any settings you will wrap it in the `wrap_table()`
function. Here we force the table headers inside the panel region and sets the
sizing of the space the table occupies based on the actual size of the table
(note, if you are using `wrap_table()` you can pass in a data.frame directly)

```{r}
p1 + wrap_table(mtcars[1:10, c('mpg', 'disp')], panel = "full", space = "fixed")
```

### Adding other content
Sometimes you need to have other content than ggplot2 in your patchwork.
Standard grid grobs can be added to your plot as well:

```{r}
p1 + grid::textGrob('Some really important text')
```

Other packages might provide even more elaborate grob specifications and these
can likewise be added directly.

Now and then, it is necessary to work with plots from the graphics package
(*base graphics*). These can be added to patchwork by providing them as a
one-sided formula:

```{r}
p1 + ~plot(mtcars$mpg, mtcars$disp, main = 'Plot 2')
```

Notice that the standard alignment you'd expect when adding ggplots together no
longer works. In general, there is no way to get consistent alignment between
ggplots and base graphics, but experiment with the different `par()` settings
until you get something that works for your particular use-case. The
[ggplotify](https://cran.r-project.org/package=ggplotify/vignettes/ggplotify.html)
package provides even more functionality for converting different graphics to
grobs so if the standard formula interface in patchwork doesn't work for you, do
check that out.

The workhorse underneath the ability to add non-ggplot objects to a patchwork is
`wrap_elements()` which is called implicitly when adding non-ggplot objects. To
get a bit more control over how your object is added, wrap the object directly
in `wrap_elements()`. Here you can define if the object should be aligned to the
full area, or to the plot area. Combining that with setting margins to zero and
not clipping the grob, you can almost get a perfect alignment:

```{r, fig.show='hold'}
old_par <- par(mar = c(0, 2, 0, 0), bg = NA)
p1 + wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE)
par(old_par)
```

An interesting side effect of this setup is that it is possible to add labels
and styling to a wrapped element (though most theme settings will be ignored).
All in all you can come pretty close to an aligned base plot, but this will
always be a bit fiddly and ad hoc. In general it is simply recommended to use
ggplot2 if at all possible.

```{r, fig.show='hold'}
old_par <- par(mar = c(0, 0, 0, 0), mgp = c(1, 0.25, 0),
               bg = NA, cex.axis = 0.75, las = 1, tcl = -0.25)
p1 +
  wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE) +
  ggtitle('Plot 2') +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 35))
par(old_par)
```

Another use case for  `wrap_elements()` is when you need the first plot to not
be a ggplot. Patchwork is not able to change the `+` behavior for miscellaneous
objects, and so, if they appear as the first element they must be wrapped in
an object that patchwork understands:

```{r}
# This won't do anything
grid::textGrob('Text on left side') + p1
```
```{r}
# This will work
wrap_elements(grid::textGrob('Text on left side')) + p1
```

### Stacking and packing
The `+` operator simply combines plots without telling patchwork anything about
the desired layout. The layout, unless changed with `plot_layout()` (See the
[Controlling Layout](layout.html) guide), will simply be a grid with enough rows
and columns to contain the number of plots, while being as square as possible.
For the special case of putting plots besides each other or on top of each
other patchwork provides 2 shortcut operators. `|` will place plots next to each
other while `/` will place them on top of each other.

```{r}
p1 / p2
```

```{r}
p1 | p2
```

For up to 3 plots `|` will behave just as `+` but using `|` will communicate the
intend of the layout better. Be aware that mixing operators will put you under
the control of the operator precedence rule (e.g. `/` will be evaluated before
`+`). Because of this it is always a good idea to put sub-assemblies within
braces to avoid any surprises:

```{r}
p1 / (p2 | p3)
```

### Functional assembly
While using the different operators provides a clean API for assembly, it is
less suited for situations where you are not in control of the plot-generation
process. If you are handed a list of plot objects and want to assemble them to
a patchwork it is a bit clumsy using the `+` operator (but doable with a loop or
`Reduce()`). patchwork provides `wrap_plots()` for a more functional approach to
assembly. It takes either separate plots or a list of them and adds them to a
single patchwork. On top of that it also accepts the same arguments as
`plot_layout()` (see the [Controlling Layouts](layout.html) guide) so it can be
used as a single solution for most assembly needs:

```{r}
wrap_plots(p1, p2, p3, p4)
```

## Nesting the left-hand side
As plots will always be added to the patchwork on the left-hand side, it is not
possible to nest the left-hand side beside the right-hand side with the standard
operators shown above. This can lead to surprising behavior:

```{r}
patch <- p1 + p2
p3 + patch
```

will nest the right-hand side, while the same is not true for the left-hand side
below

```{r}
patch + p3
```

This behavior is necessary to allow patchworks to be build up gradually, but
will get in the way if the left-hand side have to nested. patchwork provides the
`-` operator for this exact situation. It should conceptually be understood as a
hyphen and not a minus, in that it keeps each side from each other (it puts both
on the same nesting level):

```{r}
patch - p3
```

An alternative to `-` is to use `merge()` on the left hand side which does the
same thing. Using this approach it is possible to use it in combination with `/`
and `|` which is not an option when using `-`.

Yet another way to solve this is with `wrap_plots()`, which will put all its
input on the same nesting level, irrespective of order:

```{r}
wrap_plots(patch, p3)
```

## Modifying patches
When creating a patchwork, the resulting object remain a ggplot object
referencing the last added plot. This means that you can continue to add objects
such as geoms, scales, etc. to it as you would a normal ggplot:

```{r}
p1 + p2 + geom_jitter(aes(gear, disp))
```

If you need to modify another patch of the patchwork you can access and/or
modify it with double-bracket indexing. This is useful if you work with a
function that returns a patchwork and you want to modify one of the subplots:

```{r}
patchwork <- p1 + p2
patchwork[[1]] <- patchwork[[1]] + theme_minimal()
patchwork
```

### Modifying everything
Often, especially when it comes to theming, you want to modify everything at
once. patchwork provides two additional operators that facilitates this. `&`
will add the element to all subplots in the patchwork, and `*` will add the
element to all the subplots in the current nesting level. As with `|` and `/`,
be aware that operator precedence must be kept in mind.

```{r}
patchwork <- p3 / (p1 | p2)
patchwork & theme_minimal()
```
```{r}
patchwork * theme_minimal()
```

## Want more?
This is everything there is to know about combining and modifying patches in a
patchwork. Be sure to check out the other guides for more about controlling
[layouts](layout.html) and [annotations](annotation.html).
