---
title: "Alignment Across Multiple Pages"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Alignment Across Multiple Pages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patchwork)
```

While patchwork is mainly a package for creating plot compositions, it also 
contains infrastructure for a related, but distinct task: Aligning plots across
pages.

Consider the task where you are preparing a slideshow containing a range of 
figures. Optimally you'd want plots that don't *"jump around"* as you switch 
between slides, but different axis text, legends, strips, etc. can make that 
difficult. We'll illustrate this with our standard plots:

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

Notice how `p1` and `p3` doesn't align when rendered at the same size:

```{r, animation.hook='gifski'}
plot(p1)
plot(p3)
```

We can make one of the plots align with the other by extracting the dimensions
from one plot, and applying it to the other:

```{r, animation.hook='gifski'}
p3_dims <- get_dim(p3)
p1_aligned <- set_dim(p1, p3_dims)
plot(p1_aligned)
plot(p3)
```

A nice thing about the implementation is that aligned plots retain the ggplot 
nature after the dimensions have been applied, and can thus continue to be added to:

```{r}
p1_aligned + theme_minimal()
```

Often we don't have a single plot that everything else should be aligned to. 
Notice the havoc we wreck when we try to align `p4` to `p3`:

```{r}
set_dim(p4, p3_dims)
```

Because `p3` doesn't have any strips the title on `p4` will be placed too close 
to them in order to be aligned with the title of `p3`. What we want is to 
extract some dimensions that are large enough to accommodate all the different 
features in a set of plots. We can do just that with `get_max_dim()`:

```{r}
max_dims <- get_max_dim(p1, p2, p3, p4)
set_dim(p4, max_dims)
```

We can now see that the title in the other plots have been raised to make room
for the strips present in `p4`:

```{r}
set_dim(p2, max_dims)
```

Usually you'll have a list of plots and simply want to align all of these 
before rendering them to a file. All of the operations above have been wrapped
in the `align_plots()` function:

```{r, animation.hook='gifski'}
plots_aligned <- align_plots(p1, p2, p3, p4)
for (p in plots_aligned) {
  plot(p)
}
```

## Want more?
Multipage alignment is just a small part of patchwork. Check out the other 
guides to learn about [assembling plots](assembly.html), 
[specifying layouts](layout.html), and 
[annotating the results](annotation.html).
