---
title: "Plot Assembly"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Plot assembly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patchwork)
```

Before plots can be layed out, they have to be assembled. Arguably one of 
patchworks biggest selling point is that it expands on the use of `+` in ggplot2
to allow plots to be added together and composed, creating a natural extension
of the ggplot2 API. There is more to it than that, and this tutorial will teach
you all about the different operators and functions available for plot assembly.

As always, we start with a few well example plots:

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

## Adding plots to the patchwork
At this point, it shouldn't come as a surprise that you can use `+` to add plots
together to form a patchwork. If so, I'd suggest you start out with the 
*Getting Started* guide, and come back once you've gone through that. But, just
to recap, you can add plots together, like so:

```{r}
p1 + p2
```

using this approach, it is possible to assemble a number of plots, but that is
not the only thing that can be added. Other patchworks can be added, and will 
create a new nested patchwork:

```{r}
patch <- p1 + p2
p3 + patch
```

### Adding non-ggplot content
Sometimes you need to have other content than ggplot2 in your composition. 
Standard grid grobs can be added to your plot as well:

```{r}
p1 + grid::textGrob('Some really important text')
```

Other packages provide even more complex grobs to add, e.g. `tableGrob` from 
gridExtra:

```{r}
p1 + gridExtra::tableGrob(mtcars[1:10, c('mpg', 'disp')])
```

Now and then, it is necessary to work with graphics from the graphics package 
(*base graphics*). These can be added to patchwork by providing them as a
one-sided formula:

```{r}
p1 + ~plot(mtcars$mpg, mtcars$disp, main = 'Plot 2')
```

Notice that the standard alignment you'd expect when adding ggplots together no 
longer works. In general, there is no way to get consistent alignment between 
ggplots and base graphics, but experiment with the different `par()` settings
until you get something that works for your particular use-case. The 
[ggplotify](https://cran.r-project.org/package=ggplotify/vignettes/ggplotify.html)
package provides even more functionality for converting different graphics to 
grobs so if the standard formula interface in patchwork doesn't work for you, do 
check it out.

The workhorse underneath the ability to add non-ggplot objects to a patchwork is
`wrap_elements()` which is called implicitly when adding non-ggplot objects. To
get a bit more control over how your object is added, wrap the object directly 
in `wrap_elements()`. Here you can define if the object should be aligned to the
full area, or to the plot area. Combining that with setting margins to zero and 
not clipping the grob, you can almost get a perfect alignment:

```{r, fig.show='hold'}
old_par <- par(mar = c(0, 2, 0, 0), bg = NA)
p1 + wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE)
par(old_par)
```

An interesting side effect of this setup is that it is possible to add labels 
and styling to a wrapped element (though most theme settings will be ignored). 
All in all you can come pretty close to an aligned base plot, but this will 
always be a bit fidly and ad hoc. In general it is simply recommended to use
ggplot2 if at all possible.

```{r, fig.show='hold'}
old_par <- par(mar = c(0, 0, 0, 0), mgp = c(1, 0.25, 0), 
               bg = NA, cex.axis = 0.75, las = 1, tcl = -0.25)
p1 + 
  wrap_elements(panel = ~plot(mtcars$mpg, mtcars$disp), clip = FALSE) + 
  ggtitle('Plot 2') + 
  theme(plot.margin = margin(5.5, 5.5, 5.5, 35))
par(old_par)
```

Another use case for  `wrap_elements()` is when you need the first plot to not
be a ggplot. Patchwork is not able to change the `+` behaviour for miscellaneous 
objects, and so, if they appear as the first element though must be wrapped in 
an object patchwork understands:

```{r}
# This won't do anything
grid::textGrob('Text on left side') + p1
```
```{r}
# This will work
wrap_elements(grid::textGrob('Text on left side')) + p1
```

### Stacking and packing
The `+` operator simply combines plots without telling patchwork anything about
the desired layout. The layout, unless changed with `plot_layout()` (See the 
*Controlling Layout* vignette), will simply be a grid with enoug rows and 
columns to contain the number of plots, and being as square as possible. For the
special cases of putting plots besides each other or on top of each other 
patchwork provides 2 shortcut operators. `|` will place plots next to each other
while `/` will place them on top of each other.

```{r}
p1 / p2
```

```{r}
p1 | p2
```

For up to 3 plots `|` will behave just as `+` but using `|` will communicate the
intend of the layout better. Be aware that mixing operators will put you under 
the control of the operator precedence rule (e.g. `/` will be evaluated before 
`+`). Because of this it is always a good idea to put sub-assemblies within 
braces to avoid any surprises

```{r}
p1 / (p2 | p3)
```

## Modifying patches
When creating a patchwork, the resulting object remain a ggplot object 
referencing the last added plot. This means that you can continue to add ggplot
objects to it as you would a normal object:

```{r}
p1 + p2 + geom_jitter(aes(gear, disp))
```

if you need to modify another patch of the patchwork you can access and/or 
modify it with double-bracket indexing. This is useful if you work with a 
function that returns a patchwork and you want to modify one of the subplots:

```{r}
patchwork <- p1 + p2
patchwork[[1]] <- patchwork[[1]] + theme_minimal()
patchwork
```

### Modifying everything
Often, especially when it comes to theming, you want to modify everything at 
once. patchwork provides two additional operators that facilitates this. `&` 
will add the element to all subplots in the patchwork, and `*` will add the 
element to all the subplots in the current nesting level. As with `|` and `/`,
be aware that operator precedence must be kept in mind.

```{r}
patchwork <- p3 / (p1 | p2)
patchwork & theme_minimal()
```
```{r}
patchwork * theme_minimal()
```

## Want more?
This is everything there is to know about combining and modifying patches in a
patchwork. Be sure to check out the other tutorials for more about controlling 
layouts and annotations.
