---
title: "Controlling Layouts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling Layouts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patchwork)
```

While quite complex compositions can be achieved using `+`, `|`, and `/`, it may
be necessary to take even more control over the layout. All of this can be
controlled using the `plot_layout()` function along with a couple of special
placeholder objects. We'll use our well-known mtcars plots to show the different
options.

```{r}
library(ggplot2)
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp)) + 
  ggtitle('Plot 1')

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear)) + 
  ggtitle('Plot 2')

p3 <- ggplot(mtcars) + 
  geom_point(aes(hp, wt, colour = mpg)) + 
  ggtitle('Plot 3')

p4 <- ggplot(mtcars) + 
  geom_bar(aes(gear)) + 
  facet_wrap(~cyl) + 
  ggtitle('Plot 4')
```

## Adding an empty area
Sometimes all that is needed is to have an empty area in between plots. This can
be done by adding a `plot_spacer()`. It will occupy a grid cell in the same way
a plot without any outer elements (titles, ticks, strips, etc.):

```{r}
p1 + plot_spacer() + p2 + plot_spacer() + p3 + plot_spacer()
```

It is important to understand that the added area only corresponds to the size
of a plot panel. This means that spacers in separate nesting levels may have
different dimensions:

```{r}
(p1 + plot_spacer() + p2) / (plot_spacer() + p3 + plot_spacer())
```

## Controlling the grid
If nothing is given, patchwork will try to make a grid as square as possible, 
erring to the side of a horizontal grid if a square is not possible (it uses the
same heuristic as `facet_wrap()`). Further, each column and row in the grid will 
take up the same space. Both of these can be controlled with `plot_layout()`

```{r}
p1 + p2 + p3 + p4 + 
  plot_layout(ncol = 3)
```

```{r}
p1 + p2 + p3 + p4 + 
  plot_layout(widths = c(2, 1))
```

when grid sizes are given as numerics, it will define the relative sizing of the
panels. In the above, the panel area of the first column is twice that of the 
second column. It is also possible to supply a unit vector instead:

```{r}
p1 + p2 + p3 + p4 + 
  plot_layout(widths = c(2, 1), heights = unit(c(5, 1), c('cm', 'null')))
```

In the last example the first row will always occupy 5cm, while the second will
expand to fill the remaining area.

It is important to remember that sizing only affects the plotting region 
(*panel area*). If a plot have e.g. very wide y-axis text it will not be 
penalized and get a smaller overall plotting region.

## Moving beyond the grid
Earlier, when we've wanted to create non-grid compositions, we've used nesting.
While this is often enough, you end up loosing the alignement between subplots 
from different nested areas. An alternative is to define a layout design to fill
the plots into. Such a design can be defined in two different ways. The easiest
is to use a textual representation:

```{r}
layout <- "
##BBBB
AACCDD
##CCDD
"
p1 + p2 + p3 + p4 + 
  plot_layout(design = layout)
```

When using the textual representation it is your responsibility to make sure 
that each area is rectangular. The only exception is `#` which denotes empty 
areas and can thus be of any shape.

A more programmatic approach is to build up the layout using the `area()`
constructor. It is a bit more verbose but easier to program with. Further, this
allows you to overlay plots.

```{r}
layout <- c(
  area(t = 2, l = 1, b = 5, r = 4),
  area(t = 1, l = 3, b = 3, r = 5)
)
p1 + p2 + 
  plot_layout(design = layout)
```

The design specification can of course also be used with `widths` and `heights`
to specify the size of the columns and rows in the design.

# Controlling guides
Plots often have guides to help viewers deduce the aesthetic mapping. When 
composing plots we need to think about what to do about these. The default 
behaviour is to leave these alone and simply let them follow the plot around. 
Examples of this can be seen above where the colour guide is always positioned
beside Plot 3. Such behaviour is fine if the purpose is simply to position a 
bunch of otherwise unrelated plots next to each other. If the purpose of the
composition is to create a new tightly coupled visualisation, the presence of
guides in between the plots can be jarring, though. The `plot_layout()` function
provides a `guides` argument that controls how guides should be treated. To show
how it works, let's see what happens if we set it to `'collect'`:

```{r}
p1 + p2 + p3 + p4 +
  plot_layout(guides = 'collect')
```

We can see that the guide has been hoisted up and placed besides all the plots.
The alternative to `'collect'` is `'keep'`, which makes sure that guides are
kept next to their plot. The default value is `'auto'`, which is sort of in 
between. It will not collect guides, but if the patchwork is nested inside 
another patchwork and that patchwork collect guides, it is allowed to hoist them
up. To see this in effect, compare the two plots

```{r}
((p2 / p3 + plot_layout(guides = 'auto')) | p1) + plot_layout(guides = 'collect')
```

```{r}
((p2 / p3 + plot_layout(guides = 'keep')) | p1) + plot_layout(guides = 'collect')
```

The guide collection has another trick up its sleeve. If multiple plots provide
the same guide, you don't want to have it show up multiple times. When 
collecting guides it will compare them with each other and remove duplicates.

```{r}
p1a <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp, colour = mpg, size = wt)) + 
  ggtitle('Plot 1a')

p1a | (p2 / p3)
```

```{r}
(p1a | (p2 / p3)) + plot_layout(guides = 'collect')
```

Guides are compared by their graphical representation, not by their declarative
specification. This means that different theming among plots may mean that two
guides showing the same information is not merged.

Now and then you end up with an empty area in you grid. Instead of leaving it 
empty, you can specify it as a place to put the collected guides, using the
`guide_area()` placeholder. It works much the same as `plot_spacer()`, and if
guides are not collected it will do exactly the same. But if guides are 
collected they will be placed there instead of where the theme tells it to.

```{r}
p1 + p2 + p3 + guide_area() + 
  plot_layout(guides = 'collect')
```

Guide areas are only accessible on the same nesting level. If a nested patchwork
has a guide area it will not be possible to place guides collected at a higher 
level there.

## Want more?
This is all for layouts. THere is still more to patchwork. Explore the other
vignettes to learn about annotations, special operators and elements.
